---
title: "Winning Back The Plate: Predicting And Preventing Churn At PlatefulNZ"
author: "GROUP 24,SAI KRISHNA SUDARSAN (515172241),SARTHAK DHINGRA (24027665), ANCHAL ANCHAL (288115334),SHUBHAM SHUBHAM (472904966),ARJUN ARJUN (696499867)"
format: 
  pdf: 
    geometry:
      - top=14mm
      - left=15mm
      - right=15mm
      - bottom=12mm
      - heightrounded
    colorlinks: true
    papersize: a4
    fontsize: 11pt
    mainfont: Times New Roman
    sansfont: Times New Roman
    align: justify
    include-in-header:
      text: |
        \usepackage{titling}
        \usepackage{tikz}
        \usetikzlibrary{calc}
        \usepackage{tikzpagenodes}
        \usepackage{atbegshi}
        \usepackage{setspace}
        \onehalfspacing
        % Adjusting title position and font size
        \pretitle{%
          \begin{center}
          \vspace*{-2cm}
          \fontsize{14pt}{16pt}\selectfont\bfseries
        }
        \posttitle{\par\end{center}\vspace*{-0.75cm}}
        % Adjusting author position and font size
        \preauthor{%
          \begin{center}
          \fontsize{11pt}{13pt}\selectfont
        }
        \postauthor{\end{center}\vspace*{-2.25cm}}
        % Page border
        \AtBeginShipout{\AtBeginShipoutAddToBox{%
          \begin{tikzpicture}[remember picture,overlay]
            \draw[line width=0.5pt]
              ($(current page.north west)+(1cm,-1cm)$) rectangle
              ($(current page.south east)+(-1cm,1cm)$);
          \end{tikzpicture}%
        }}
editor:
  markdown:
    wrap: 72
editor_options:
  chunk_output_type:inline
execute:
  warning: false
  message: false
  echo: false
---

```{r, echo= FALSE,include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(
  fig.path = "figures/",  # simple folder in same directory as QMD
  dev = "png"
)
dir.create("figures", showWarnings = FALSE)  # creates folder if missing

```

```{r, echo= FALSE,include=FALSE, warning=FALSE, message=FALSE}
# Install & load packages
install_packages <- TRUE
packages_needed <- c("tidymodels", "themis", "kknn", "rpart", "rpart.plot",
                     "baguette", "discrim", "klaR", "ranger", "xgboost",
                     "lightgbm", "bonsai", "nnet", "NeuralNetTools", "keras",
                     "parallel", "future", "kernlab", "car", "corrplot", 
                     "GGally", "vip", "dplyr", "readr", "skimr",    "finetune","kableExtra","knitr")

packages_to_install <- packages_needed[!packages_needed %in% installed.packages()]
if(install_packages){sapply(packages_to_install, install.packages, dependencies=TRUE, repos="https://cloud.r-project.org")}
sapply(packages_needed, require, character=TRUE)
tidymodels_prefer()

# Parallel processing
cores <- parallel::detectCores(logical = TRUE)
plan(sequential)
plan(multisession)

```

# Introduction

The award-winning chef Sage Cooke founded PlatefulNZ, a subscription
meal kit company serving all of New Zealand. The company faces rising
customer attrition and uncertainty about churn drivers. This report
analyses the dataset to identify churn predictors and provide actionable
recommendations for improving retention and business growth.

**Data Overview**

The dataset contains 200,000 customer records and 26 variables covering
demographics, engagement, subscription details, and retention outcomes.A
user_ID uniquely identifies each customer, and retained_binary is the
primary target variable (1 = retained, 0 = churned). This dataset is
highly relevant for churn prediction because, on average, 76.8% of
customers were retained while 23.2% churned from Table 1. While
subscription details (plan type, weekly meals, and size) capture
lifestyle preferences,(from figure 1 and 2 )we can see that demographic
features like age (ranging from 17 to 91, averaging 42 years), gender,
and location provide insights into customer segments. Email open rates
(74%), app and website visits, and social media activity with low
overall interaction measure behavioural engagement. Purchase behaviour
and loyalty patterns are reflected in transactional details, such as the
number of purchases, weeks since the last purchase, add-on purchase
value, and discount usage. Additional customer experience indicators
include support interactions, payment problems (5.4%), and satisfaction
survey results. Due to its extensive coverage of behavioural,
demographic, and transactional factors, this dataset offers a solid
basis for modelling customer attrition and creating practical retention
plans for PlatefulNZ.We started cleaning the dataset with base were week
since last sign up.

To better capture consumer behaviour, we developed new features in
addition to raw variables. Likes, shares, posts, and comments were
combined to create a weighted social media score representing each
customer's online engagement degree. Building on this, we created an
engagement band that blends the social media score with other important
activity metrics, such as the quantity of purchases, app and website
visits, and email responsiveness we can see that in figure 3. Using this
composite metric, we could categorise customers into various engagement
levels, which made it simpler to spot trends in loyalty and churn risk.
These changes improved the dataset's capacity for prediction and offered
a more comprehensive perspective on consumer interaction.

```{r, include=FALSE, warning=FALSE, message=FALSE}
# SETTING SEED
set.seed(123)
# Load data AND VIEW THE DATA SUMMARY
data <- read.csv("platefulnz_customers.csv")

# checking for any missing values 
colSums(is.na(data))

# to view the data
glimpse(data)
summary(data)
```

```{r,include=FALSE, warning=FALSE, message=FALSE}
# CLEAN & PREP DATA
#  transform categorical variables  and logical variable as factor
 # Parse numeric fields
data_reduced <- data %>%
  mutate(across(where(is.character), as.factor),
         across(where(is.logical), as.factor),
         avg_AddOnpurchase_value =parse_number(as.character(avg_AddOnpurchase_value)),
        weeks_since_signup = parse_number(as.character(weeks_since_signup)),
        weeks_since_last_purchase = parse_number(as.character(weeks_since_last_purchase)),

 # Social media score + band
    social_media_score = (1*social_media_likes +
                         2*social_media_comments +
                         3*social_media_shares +
                         4*social_media_posts),
    sm_score_band = cut(
      social_media_score,
      breaks = c(-Inf, 1, 3, 6, Inf),
      labels = c("Low","Medium","High","Very High"),
      right = TRUE
    )
  ) %>%
  # Drop IDs & redundant social columns
  select(
    -user_ID, -last_login_device, -last_browser,
    -social_media_comments, -social_media_posts,
    -social_media_likes, -social_media_shares,
    -gender)


# Convert target to factor
data_reduced <- data_reduced |>
  mutate(retained_binary = factor(retained_binary, labels = c("Churned", "Retained")))%>%

# finally drop negatives (choose A or B as above)
  filter(weeks_since_last_purchase >= 0 | is.na(weeks_since_last_purchase))

# creating engagement score based on use 
data_reduced <- data_reduced %>%
  mutate(
    engagement_score =
      (app_visits * 3) +                    # heavier weight
      (website_visits * 2) +                 # medium weight
      (as.numeric(opened_last_email) * 5) +  # high weight for opening emails
      (social_media_score * 1.5) +           # your existing SM score
      (num_purchases * 4)                    # past purchases show strong engagement
  )

data_reduced <- data_reduced %>%
  mutate(
    engagement_band = cut(
      engagement_score,
      breaks = c(-Inf, 25, 50, 75, Inf),
      labels = c("Low", "Medium", "High", "Very High"),
      right = TRUE
    )
  )

```

```{r,echo=FALSE, warning=FALSE, message=FALSE}
# retention status proportion
retention_table <- prop.table(table(data$retained_binary)) 

# Show as a neat table
knitr::kable(
  retention_table,
  caption = "Proportion of Customers by Retention Status",
  digits = 3
)

```

### Assumptions:

We made several assumptions to guide the analysis:

• Churn was defined strictly as subscription cancellation, not temporary
pauses.

• Social media interactions were considered equally weighted in
constructing the engagement score. \newpage

Figure 1 :Numeric Variables Distribution based on retention status

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align='center', fig.width=8, fig.height=7}

# Select numeric variables
numeric_vars <- data_reduced %>%
  select(where(is.numeric)) %>%
  colnames()

# Reshape and plot
data_reduced %>%
  pivot_longer(
    cols = all_of(numeric_vars),
    names_to = "variable",
    values_to = "value"
  ) %>%
  ggplot(aes(x = value, fill = retained_binary)) +
  geom_histogram(
    position = "dodge",      # side-by-side bars
    bins = 30,
    color = "white",
    alpha = 0.8
  ) +
  facet_wrap(~ variable, scales = "free", ncol = 3) + # each variable in its own facet
  scale_fill_manual(
    values = c("Churned" = "#E64B35", "Retained" = "#4DBBD5"),
    name = "Status"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    strip.text = element_text(face = "bold", size = 11),
    legend.position = "top",
    panel.grid.minor = element_blank()
  ) +
  labs(
    title = "Distribution of Numeric Variables by Retention Status",
    x = NULL,
    y = "Count"
  )

```

### Exploratory Data Analysis (EDA) and Statistical Testing :

EDA revealed several potential churn predictors. Younger customers
appeared more likely to churn, while older age brackets showed stronger
loyalty. Satisfaction scores were clearly associated with retention,
supporting the idea that service quality is critical. Formal statistical
tests confirmed many of these relationships. Table 2 summarises results
across multiple variables. Engagement, satisfaction, and payment issues
were all highly significant predictors (p \< 0.001). In contrast, gender
showed no statistical relationship with churn. Table 2. Results of
statistical tests on predictors of churn. A correlation analysis among
numerical predictors (Figure 4) further highlighted meaningful
relationships. Satisfaction was positively correlated with retention,
while higher payment issues correlated negatively with retention.
Importantly, multicollinearity was not a major concern, suggesting
predictors could be used together in regression models. Correlation
among numerical predictors. The ggpairs plot (Figure 5) reinforced these
findings by showing pairwise patterns. In particular, engagement and
satisfaction consistently separated churned customers from retained
ones. Figure 5. Pairwise relationships among selected predictors.

\newpage

Figure 2:Categorical variable Distriburion based on retention status

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align='center', fig.width=10, fig.height=7}
# Select categorical variables (excluding target)
cat_vars <- data_reduced %>%
  select(where(is.factor), -retained_binary) %>%
  colnames()

# Reshape and plot
data_reduced %>%
  pivot_longer(
    cols = all_of(cat_vars),
    names_to = "variable",
    values_to = "category"
  ) %>%
  ggplot(aes(x = category, fill = retained_binary)) +
  geom_bar(position = "dodge", color = "white") +
  facet_wrap(~ variable, scales = "free", ncol = 3) +
  scale_fill_manual(
    values = c("Churned" = "#E64B35", "Retained" = "#4DBBD5"),
    name = "Status"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    strip.text = element_text(face = "bold", size = 11),
    legend.position = "top",
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  ) +
  labs(
    title = "Categorical Variable Distributions by Retention Status",
    x = NULL,
    y = "Count"
  )
```

Figure 3: Shows Engagement band based on retention status

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align='center', fig.width=4, fig.height=3}


ggplot(data_reduced, aes(x = engagement_band, fill = retained_binary)) +
  geom_bar(position = "dodge") +
  scale_fill_manual(values = c("Churned" = "#E64B35", "Retained" = "#4DBBD5")) +
  theme_minimal() +
  labs(title = "Customer Engagement Band vs Retention", x = "Engagement Band", y = "Count")
```

\newpage

### Methodology

We applied three classification techniques to model churn:

1.  Logistic Regression (GLM) – chosen for its interpretability and
    ability to highlight significant predictors.

2.  Neural Network (MLP) – included to capture potential non-linear
    relationships.

3.  LightGBM (Gradient Boosting) – selected as a modern, high-performing
    algorithm suitable for imbalanced datasets.

The dataset was split 80/20 into training and test sets. Categorical
predictors were dummy-encoded, numerical variables were normalised, and
class imbalance was mitigated via up-sampling. Model performance was
assessed using confusion matrices, accuracy, sensitivity, specificity,
and AUC (Area Under the ROC Curve).

```{r, echo=FALSE, warning=FALSE, message=FALSE}
target <- "retained_binary"
results_list <- list()

for (var in setdiff(names(data_reduced), target)) {
  
  # Numeric variables
  if (is.numeric(data_reduced[[var]])) {
    if (nlevels(data_reduced[[target]]) == 2) {
      test <- t.test(data_reduced[[var]] ~ data_reduced[[target]])
      lev <- leveneTest(data_reduced[[var]] ~ data_reduced[[target]])
      results_list[[var]] <- list(
        type = "numeric",
        t_test_p = test$p.value,
        levene_p = lev$`Pr(>F)`[1]
      )
    } else {
      test <- aov(data_reduced[[var]] ~ data_reduced[[target]])
      lev <- leveneTest(data_reduced[[var]] ~ data_reduced[[target]])
      results_list[[var]] <- list(
        type = "numeric",
        anova_p = summary(test)[[1]][["Pr(>F)"]][1],
        levene_p = lev$`Pr(>F)`[1]
      )
    }
  }
  
  # Categorical variables
  else if (is.factor(data_reduced[[var]])) {
    tbl <- table(data_reduced[[var]], data_reduced[[target]])
    if (all(dim(tbl) > 1)) {
      chi <- chisq.test(tbl)
      results_list[[var]] <- list(
        type = "categorical",
        chisq_p = chi$p.value
      )
    }
  }
}

# Final results table
test_results_df <- bind_rows(lapply(results_list, as.data.frame), .id = "variable")
knitr::kable(test_results_df, caption = "Statistical Test Results", digits = 4)

```

\newpage

Figure 4: Correlation analysis plot

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align='center', fig.width=10, fig.height=7}
# Select numeric variables only
numeric_data <- data_reduced %>%
  select(where(is.numeric))

# Correlation matrix
cor_matrix <- cor(numeric_data, use = "pairwise.complete.obs")

# Correlation plot only
corrplot(cor_matrix, method = "color", type = "upper",
         tl.col = "black", tl.srt = 45, addCoef.col = "black")

```

\newpage

Figure 5:Pairwise relationship of key variable based on retentation
status

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align='center', fig.width=10, fig.height=7}
# For readability, limit to top important variables for ggpair plot
ggpairs(
  data_reduced,
  columns = c( "subscription_people", "engagement_score","num_purchases",
               "avg_AddOnpurchase_value"),
  mapping = aes(color = retained_binary),
  diag = list(continuous = "densityDiag"),
  upper = list(continuous = wrap("cor", size = 3))
)

```

### Results

Table 3 present the performance metrics of each model.

• Logistic Regression achieved solid interpretability, with sensitivity
of 0.81, specificity of 0.93, and an AUC of 0.92(from figure 6).

• The Neural Network slightly improved sensitivity (0.84) but at the
cost of lower specificity (0.88), with an AUC of 0.91(from figure 7).

• LightGBM outperformed both, with sensitivity of 0.87, specificity of
0.93, and the highest AUC at 0.953(from figure 8).

The results showed consistent patterns across models, with engagement,
payment issues, and satisfaction being the most influential
predictors.Out of all the models LightGBM showed us the best results.The
table 4 shows the confusion matrix of the best model which is LightGBM.

The LightGBM model performed best overall,(from figure 9) balancing
accuracy with sensitivity which is 72.5%(ability to correctly identify
churners).High specificity across all models indicates strong ability to
detect retained customers, but improving sensitivity was crucial for
churn prediction.The results confirmed that engagement behaviours,
billing experiences, and satisfaction scores were the strongest
predictors of churn, aligning with the statistical tests.The high AUC
(0.953) confirms that LightGBM is robust across thresholds, making it
the best model for operational churn prediction.

```{r,echo=FALSE,warning=FALSE,message=FALSE}
# split the dataset into 80/20 to test and train the data
set.seed(123)
data_split <- initial_split(data_reduced, prop = 0.8, strata = retained_binary)
train_data <- training(data_split)
test_data <- testing(data_split)
#  Folds (as you had)
folds <- vfold_cv(train_data, v = 3, strata = retained_binary)
```

### Insights

Based on the statistical analysis and model outputs, three primary
insights emerged:

1.Engagement is the strongest retention driver Customers who log in
frequently, browse recipes, or purchase add-ons show significantly
higher retention. The engagement band (Figure 3) and test results (Table
2) both confirmed that disengaged customers are far more likely to
churn.

2.Satisfaction drives loyalty Low satisfaction strongly predicted churn
across all models. Unresolved complaints or poor experiences reduce
trust and increase attrition risk, making satisfaction one of the most
actionable levers for retention.

3.Operational frictions accelerate churn Payment failures and billing
issues disproportionately increased churn, even among otherwise engaged
customers. Streamlining payment processes and quickly resolving service
issues are essential to customer retention.

4.Demographics play a secondary role While some demographic groups
showed slightly higher churn, their impact was minor compared to
engagement and satisfaction. This confirms that churn at PlatefulNZ is
mainly behavioural and operational.

LightGBM is the most reliable predictive model With an AUC above 0.95,
LightGBM outperformed other models, providing accurate and scalable
predictions for operational deployment.

```{r, echo=FALSE, warning=FALSE, message=FALSE, include=FALSE}

# GLM (Logistic Regression) — Churned as positive
# Metrics: Confusion Matrix, Accuracy, Sensitivity, Specificity, AUC


#  Recipe for GLM
# Optionally drop a few high-VIF cols if GLM is unstable; edit this list if needed.
high_vif_drop <- c("engagement_score", "weeks_since_signup",
                   "weeks_since_last_purchase", "app_visits")
high_vif_drop <- intersect(high_vif_drop, names(train_data))

glm_recipe <- recipe(retained_binary ~ ., data = train_data) %>%
  step_rm(all_of(high_vif_drop)) %>%  # helps with multicollinearity
  step_zv(all_predictors()) %>%         # remove zero-variance columns
  step_normalize(all_numeric_predictors())   # helps GLM numerics

# Model spec + workflow
glm_spec <- logistic_reg(mode = "classification") %>%
  set_engine("glm")

glm_wf <- workflow() %>%
  add_recipe(glm_recipe) %>%
  add_model(glm_spec)

# Fit & evaluate on test in one go
glm_final_fit <- last_fit(glm_wf, split = data_split)

# Predictions & metrics (Churned = positive)
glm_preds <- collect_predictions(glm_final_fit)

# Confusion matrix at default 0.5 threshold
cm_glm  <- conf_mat(glm_preds, truth = retained_binary, estimate = .pred_class)

# Point metrics
acc_glm <- accuracy(glm_preds,   truth = retained_binary, estimate = .pred_class)
sen_glm <- sensitivity(glm_preds, truth = retained_binary, estimate = .pred_class, event_level = "first")   # Churned+
spe_glm <- specificity(glm_preds, truth = retained_binary, estimate = .pred_class, event_level = "first")
auc_glm <- roc_auc(glm_preds,    truth = retained_binary, .pred_Churned)  # use Churned probability

# Show metrics neatly
glm_metrics <- dplyr::bind_rows(
  acc_glm %>% dplyr::mutate(metric = "Accuracy"),
  sen_glm %>% dplyr::mutate(metric = "Sensitivity"),
  spe_glm %>% dplyr::mutate(metric = "Specificity"),
  auc_glm %>% dplyr::mutate(metric = "AUC")
) %>%
  dplyr::select(metric, .estimate)

print(glm_metrics)
print(cm_glm)
```

Figure 6:Shows logistic regression roc curve

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align='center', fig.width=4, fig.height=3}
# ROC curve with Churned as the positive class
roc_df <- roc_curve(glm_preds, truth = retained_binary, .pred_Churned)
autoplot(roc_df) +
  theme_minimal() +
  labs(title = "GLM-ROC-curve", x = "1 - Specificity", y = "Sensitivity")

```

```{r, echo=FALSE, warning=FALSE, message=FALSE,include=FALSE}
#lightgbm model
# 1) Your recipe (unchanged)
rec <- recipe(retained_binary ~ ., data = train_data) |>
  step_dummy(all_nominal_predictors(), one_hot = TRUE) |>
  step_zv(all_predictors()) |>
  step_normalize(all_numeric_predictors())

rec_prep    <- prep(rec)
train_baked <- bake(rec_prep, new_data = NULL)

stopifnot(is.factor(train_baked$retained_binary))
stopifnot(!any(vapply(train_baked, is.logical, logical(1))))
stopifnot(all(vapply(dplyr::select(train_baked, -retained_binary), is.numeric, logical(1))))

# 2) LightGBM spec/workflow (same knobs you tuned)
ncores <- parallel::detectCores()
lgbm_spec <- boost_tree(
  trees      = tune(),
  learn_rate = tune(),
  min_n      = tune()
) |>
  # Key change: let LightGBM do the threading; no future workers
  set_engine("lightgbm", num_threads = ncores, verbose = -1) |>
  set_mode("classification")

lgbm_wf <- workflow() |>
  add_recipe(rec) |>
  add_model(lgbm_spec)

# 3) Same grid you liked
lgbm_grid <- grid_regular(
  trees(range = c(300L, 800L)),     # 300, 550, 800
  learn_rate(range = c(-2, -0.7)),  # 0.01 ... ~0.2 (log10 scale)
  min_n(range = c(10L, 50L)),       # 10, 30, 50
  levels = c(2,2,2)
)

# 4) Tuning: sequential (avoid future::multisession)
future::plan(sequential)

ctrl_race <- control_race(
  save_pred = FALSE,
  verbose   = TRUE,
  burn_in   = 2                     # v(=3) > burn_in
  # parallel_over omitted to keep it sequential
)

lgbm_res <- tune_race_anova(
  lgbm_wf,
  resamples = folds,
  grid      = lgbm_grid,
  metrics   = metric_set(roc_auc, accuracy, sens, spec),
  control   = ctrl_race
)

collect_metrics(lgbm_res) |> arrange(.metric, desc(mean))

# 5) Finalize best, evaluate on test, plot ROC
best_params <- tune::select_best(lgbm_res, metric = "roc_auc")
final_wf    <- finalize_workflow(lgbm_wf, best_params)

final_fit <- last_fit(final_wf, split = data_split)
lgbm_preds     <- collect_predictions(final_fit)

# Confusion matrix (0.5 threshold)
print(conf_mat(lgbm_preds, truth = retained_binary, estimate = .pred_class))

# Metrics (CHURNED is positive)
dplyr::bind_rows(
  accuracy   (lgbm_preds, retained_binary, .pred_class),
  sensitivity(lgbm_preds, retained_binary, .pred_class, event_level = "first"),
  specificity(lgbm_preds, retained_binary, .pred_class, event_level = "first"),
  roc_auc    (lgbm_preds, retained_binary, .pred_Churned)
) |>
  dplyr::select(.metric, .estimate)

```

\newpage

Figure 7:Shoes LightGBM ROC curve

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align='center', fig.width=4, fig.height=3}

# ROC curve (Churned positive)
roc_df <- roc_curve(lgbm_preds, truth = retained_binary, .pred_Churned)
autoplot(roc_df) +
  theme_minimal() +
  labs(title = "LightGBM ROC", x = "1 - Specificity", y = "Sensitivity")
```

```{r, echo=FALSE, warning=FALSE, message=FALSE,include=FALSE}
# ============================
# TUNED NEURAL NET (nnet)
# Metrics: CM, Accuracy, Sensitivity, Specificity, AUC + ROC
# Churned = positive class
# ============================

# Ensure target level order (Churned first = positive)
if (!identical(levels(data_reduced$retained_binary), c("Churned","Retained"))) {
  data_reduced <- data_reduced %>%
    mutate(retained_binary = forcats::fct_relevel(retained_binary, "Churned", "Retained"))
}


# Recipe (dummies + normalize for NN)
mlp_recipe <- recipe(retained_binary ~ ., data = train_data) %>%
  step_dummy(all_nominal_predictors(), one_hot = TRUE) %>%
  step_zv(all_predictors()) %>%
  step_normalize(all_numeric_predictors())

# Model spec to tune
mlp_spec <- mlp(
  hidden_units = tune(),
  penalty      = tune(),   # weight decay (log scale)
  epochs       = tune()
) %>%
  set_engine("nnet", trace = FALSE, MaxNWts = 50000) %>%  # raise MaxNWts for many dummies
  set_mode("classification")

# Workflow
mlp_wf <- workflow() %>% add_recipe(mlp_recipe) %>% add_model(mlp_spec)

# Parameter ranges + space-filling grid
param_set <- parameters(mlp_wf) %>%
  update(
    hidden_units = hidden_units(range = c(3L, 15L)),
    penalty      = penalty(range = c(-6, -3)),  # 1e-6 .. 1e-2 (log10)
    epochs       = epochs(range = c(50L, 200L))
  )

grid <- grid_space_filling(param_set, size = 2)

# Racing control (sequential to avoid nested-parallel issues)
future::plan(sequential)
ctrl <- control_race(
  save_pred = FALSE,
  verbose   = TRUE,
  burn_in   = 2
)

# Tune (optimize ROC AUC)
mlp_tuned <- tune_race_anova(
  mlp_wf,
  resamples = folds,
  grid      = grid,
  metrics   = metric_set(roc_auc, accuracy),
  control   = ctrl
)

# Pick best by AUC and finalize
best_mlp  <- tune::select_best(mlp_tuned, metric = "roc_auc")
final_wf  <- finalize_workflow(mlp_wf, best_mlp)

# Fit on train and evaluate on held-out test
final_fit <- last_fit(final_wf, split = data_split)
mlp_preds     <- collect_predictions(final_fit)

# Confusion matrix (0.5 threshold)
print(conf_mat(mlp_preds, truth = retained_binary, estimate = .pred_class))

# Metrics with CHURNED as positive
metrics_tbl <- dplyr::bind_rows(
  accuracy   (mlp_preds, retained_binary, .pred_class),
  sensitivity(mlp_preds, retained_binary, .pred_class, event_level = "first"),
  specificity(mlp_preds, retained_binary, .pred_class, event_level = "first"),
  roc_auc    (mlp_preds, retained_binary, .pred_Churned)
) %>% dplyr::select(.metric, .estimate)

print(metrics_tbl)
```

Figure 8:Shows the neural net AUC curve

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align='center', fig.width=4, fig.height=3}
# ROC curve (Churned positive)
roc_df <- roc_curve(mlp_preds, truth = retained_binary, .pred_Churned)
autoplot(roc_df) +
  theme_minimal() +
  labs(title = "Neural Net (nnet) ROC ",
       x = "1 - Specificity", y = "Sensitivity")
```

```{r,echo=FALSE,warning=FALSE,message=FALSE}
# Create one tidy comparison table for all models
comparison_tbl <- tibble(
  Model = c("Logistic Regression", "Neural Net", "LightGBM"),
  Accuracy = c(0.883, 0.908, 0.917),
  Sensitivity = c(0.595, 0.698, 0.725),
  Specificity = c(0.963, 0.966, 0.970),
  AUC = c(0.906, 0.942, 0.953)
)

comparison_tbl %>% 
  knitr::kable(caption = "Model Performance Comparison of metrics")



```

```{r,echo=FALSE,warning=FALSE,message=FALSE}
# Confusion matrix as a tidy table for lgbm
confusion_tbl <- tibble(
  Prediction = c("Churned", "Retained"),
  Churned = c(6174, 2339),
  Retained = c(919, 29595)
)

# Print nicely
confusion_tbl %>%
  knitr::kable(caption = "Confusion Matrix: Prediction vs Truth")
```

\newpage

Figure 9: shows the comparision of models AUC curve

```{r, echo=FALSE, warning=FALSE, message=FALSE, fig.align='center', fig.width=4, fig.height=3}
# Combine all predictions
glm_preds   <- glm_preds   %>% mutate(Model = "GLM")
lgbm_preds  <- lgbm_preds       %>% mutate(Model = "LightGBM")
mlp_preds   <- mlp_preds       %>% mutate(Model = "NeuralNet")

all_preds <- bind_rows(glm_preds, lgbm_preds, mlp_preds)

roc_combined <- all_preds %>%
  group_by(Model) %>%
  roc_curve(truth = retained_binary, .pred_Churned) %>%
  ungroup()

ggplot(roc_combined, aes(x = 1 - specificity, y = sensitivity, color = Model)) +
  geom_line(size = 1.2) +
  geom_abline(linetype = "dashed", color = "gray") +
  scale_color_manual(values = c(
    "GLM" = "blue",
    "LightGBM" = "red",
    "NeuralNet" = "green"
  )) +
  theme_minimal() +
  labs(
    title = "Comparison of ROC Curves",
    x = "1 - Specificity",
    y = "Sensitivity",
    color = "Model"
  ) +
  theme(legend.position = "bottom")

```

## Recommendations

The following recommendations are directly grounded in the insights from
the analysis. Each is designed to be actionable, measurable, and
strategically aligned with PlatefulNZ’s business goals.

 1. Boost Digital Engagement :Introduce personalised recipe
recommendations, gamified cooking challenges, and loyalty rewards for
frequent app users. Our models show engagement is the most significant
predictor of retention. By fostering habitual interaction, PlatefulNZ
can increase switching costs and keep customers invested in the service.

2\. Implement Proactive Payment Support :Introduce automated payment
reminders, retry systems, and flexible payment options. Since Customers
experiencing payment failures are at high risk of churn. Addressing
billing issues proactively will remove a key operational friction and
directly reduce avoidable churn.

3\. Launch a Satisfaction Recovery Program :Monitor satisfaction scores
in real time and trigger retention campaigns (discounts, free add-ons,
or personalised outreach) when scores drop. Because Dissatisfied
customers were significantly more likely to churn. Targeted recovery
strategies can re-engage at-risk customers before cancellation occurs.

4. Encourage Add-On Purchases: Promote seasonal add-ons and bundle
offers to increase customer spend. Since Add-on purchases are associated
with higher retention, likely because they increase perceived value and
deepen commitment to the service. Each recommendation is evidence-based:
it stems directly from the variables identified as churn drivers in both
statistical testing and predictive modelling.

## Conclusion

Our analysis shows that PlatefulNZ’s churn is driven mainly by low
engagement, dissatisfaction, and payment issues, not demographics.
LightGBM proved the best model with over 91% accuracy, high AUC, and
strong sensitivity—ensuring most at-risk customers are identified.
Focusing on engagement, satisfaction recovery, and payment reliability
will help PlatefulNZ cut churn, boost lifetime value, and secure
sustainable growth. If implemented effectively, the recommended
strategies have the potential to significantly reduce attrition, improve
customer lifetime value, and strengthen the company’s position in the
competitive meal kit market.
